taxize - taxonomic search and retrieval in R

### 2013-07-05 inst/doc/taxize_withcode_plos.Rnw
added and fixed references + minor changes

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming easily available on the web, but scientists need a way to access taxonomic data on the web in a programmatic fashion that's easy and reproducible. We have developed taxize, an open-source software package (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}) for the R language. taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package will facilitate open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.

@@

The R language is the dominant language used by biologists (reference), and now has over 4,500 packages on the Comprehensive R Archive Network (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. However, the key missing piece in R is the ability to get data from the internet within R. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Increasingly, data is available from the web via API's, or application programming interfaces. These are bits of code that allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further strength of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Whereas, if data are stored in an Excel file, for example, the file can only be opened in a few programs.

@@

One data source available in taxize is available in another R package that solely interact with that data source. We provide a few convenience functions that wrap functions in \emph{taxonstand}, an R interface to The Plant List \cite{theplantlist}.

@@

New data sources can be added; we may add the following sources: Wikispecies and The Tree of Life. A connection to \url{freshwaterecology.info} \cite{freshwaterecology} (a database with autecological characteristics, ecological preferences and biological traits as well as distribution patterns of more than 12.000 European freshwater organisms belonging to fish, macro-invertebrates, macrophytes, diatoms and phytoplankton) will be finished when their new API will be released. Get in touch with one of the authors if you would like any data sources added.

@@

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, functional trait ecology \cite{REF}. Yet, Many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in figure \ref{fig:phylomatic}.

@@

\subsection*{What taxa are the children of my taxon of interest?}
If you aren't a taxonomic specialist on a particular taxon you likely don't know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies \cite{wikispecies} or Encyclopedia of Life \cite{eol} to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL) \cite{col} and the Integrated Taxonomic Information System \cite{itis}. Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

@@

\subsection*{IUCN Status}
There are a number of things we can do once we have the correct taxonomic names. One thing we can do is ask about the conservation status of a species. We have provided a set of functions, \emph{iucn\_summary} and \emph{iucn\_status}, to search for species names, and extract the status information, respectively. Here, we search for the Panther and Lynx.

@@

It turns out that there are XX different unique records found. However, this doesn't mean that there are XX different genes found as the API does not provide metadata to classify genes. However, at the end of the example, we showed that you can use regular expressions via (e.g., via \emph{grep}) to search for the gene of interest.

@@

<<aggregate_abd, eval=TRUE, cache=TRUE>>=
library(vegan)
data(dune.taxon)
dat <- dune.taxon
dat$abundance <- round(rlnorm(n = nrow(dat), meanlog = 5, sdlog = 2), 0)
head( rankagg(data = dat, datacol = "abundance", rank = "Genus") )
head( rankagg(data = dat, "abundance", rank = "Family", fxn = "mean") )
@

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming increasingly available on the web, but scientists need a way to access taxonomic data on the web in a programmatic fashion that's easy and reproducible. We have developed taxize, an open-source software package (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}) for the R language. taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package will facilitate open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.

@@

The R language is the dominant language used by biologists (reference), and now has over 4,500 packages on the Comprehensive R Archive Network (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. However, the key missing piece in R is the ability to get data from the internet within R. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Increasingly, data is available from the web via application programming interfaces (API). These are bits of code that allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further strength of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Whereas, if data are stored in an Excel file, for example, the file can only be opened in a few programs.

@@

One available data source in taxize is The Plant List \cite{theplantlist}. The connection in taxize is done via the \emph{taxonstand} package \cite{Cayuela_2012} that solely interacts with that The Plant List. We provide a few convenience functions that wrap taxonstand into taxize.

@@

New data sources can be added; we may add the following sources: Wikispecies and The Tree of Life. A connection to \url{freshwaterecology.info} \cite{freshwaterecology} (a database with autecological characteristics, ecological preferences and biological traits as well as distribution patterns of more than 12.000 European freshwater organisms belonging to fish, macro-invertebrates, macrophytes, diatoms and phytoplankton) will be finished when their new API will be released.
In addition, the authors may be contacted for further suggestions of data sources to be added.

@@

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, functional trait ecology \cite{poff2006functional}. Yet, Many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in figure \ref{fig:phylomatic}.

@@

\subsection*{What taxa are the children of my taxon of interest?}
If someone is not a taxonomic specialist on a particular taxon he likely does not know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies \cite{wikispecies} or Encyclopedia of Life \cite{eol} to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL) \cite{col} and the Integrated Taxonomic Information System \cite{itis}. Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

@@

\subsection*{IUCN Status}
There are a number of things we can do once we have the correct taxonomic names. One thing we can do is ask about the conservation status of a species (IUCN Red List of Threatened Species \cite{iucn}). We have provided a set of functions, \emph{iucn\_summary} and \emph{iucn\_status}, to search for species names, and extract the status information, respectively. Here, we search for the Panther and Lynx.

@@

It turns out that there are XX different unique records found. However, this doesn't mean that there are XX different genes found as the API does not provide metadata to classify genes. However, at the end of the example, we showed that you can use regular expressions (e.g., via \emph{grep}) to search for the gene of interest.

@@

<<aggregate_abd, eval=TRUE, cache=TRUE>>=
library(vegan)
data(dune.taxon)
dat <- dune.taxon
dat$abundance <- round(rlnorm(n = nrow(dat), meanlog = 5, sdlog = 2), 0)
head(dat)
head( rankagg(data = dat, datacol = "abundance", rank = "Genus") )
@
